---
title: "Project_Relationship between Food Consumption Patterns and Nutritional Status in France"
output: pdf_document
Author: "DoHyun Seo"
date: "2023-11-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

##Library
```{r install library, message=FALSE, results='hide'}
install.packages("FactoMineR", repos = "https://cran.r-project.org")
install.packages("factoextra", repos = "https://cran.r-project.org")
install.packages("cluster", repos = "https://cran.r-project.org")
library(FactoMineR)
library(factoextra)
library(dplyr)
library(tidyverse)
library(gtsummary)
library(cluster)
library(epiDisplay)
```

## Load data
```{r load data}
inca <- read.csv("inca2_survey.csv", header = T, sep=",", dec = ".")
```

##Data Management
Recode variable
```{r bmi into binary (not overweight vs overweight)}
inca$bmi_bin <- ifelse(inca$bmiclass<3, 0, 1)
inca$bmi_bin <- factor(inca$bmi_bin, levels = c(0,1), labels=c("Not overweight","Overweight"))
```

```{r smoking into binary (smoking vs not smoking)}
inca$smoking_bin <- ifelse(inca$smoking_status<3, 1, 0)
inca$smoking_bin <- factor(inca$smoking_bin, levels = c(0,1), labels=c("Not smoking","Smoking"))
```

```{r education into binary (not university vs university)}
inca$education_bin <- ifelse(inca$education<3, 0, 1)
inca$education_bin <- factor(inca$education_bin, levels = c(0,1), labels=c("No university","University"))
```

Transform variable into factor
```{r transform variable}
inca$diet <- factor(inca$diet, levels = c(0,1), labels=c("No diet","Diet"))
inca$disease <- factor(inca$disease, levels = c(0,1), labels=c("No CD","CD"))
inca$season <- factor(inca$season, levels = c(1,2,3,4), labels=c("Winter","Spring","Summer","Autumn"))
inca$age_categories <- factor(inca$age_categories, levels = c(1,2,3), labels=c("young adult","middle aged","elderly"))
inca$ipaqnx <- factor(inca$ipaqnx, levels = c(1,2), labels=c("Normal PA","Sedentary"))
inca$supplements <- factor(inca$supplements, levels = c(0,1), labels=c("No supplements","Supplements"))
inca$income <- factor(inca$income, levels = c(1,2,3,4), labels=c("Low SES","Lower middle SES","Higher middle SES","High SES"))
inca$household_4c <- factor(inca$household_4c, levels = c(1,2,3,4), labels=c("Single","Single parent","Couple no children","Couple with children"))
```


```{r clean data}
inca_clean <- na.omit(inca)
summary(inca_clean)
```

#Perform PCA
Create subset
```{r create subset}
inca_pca <- inca_clean[,-c(1,3,4,7,8,51,52,55,56)]
```

```{r pca}
res.pca <- PCA(inca_pca, scale.unit = T, ncp = 14, quali.sup= c(1,2,3,46,47), graph =F)
```

```{r choose dimensions}
#Check eigen value
res.pca$eig
barplot(res.pca$eig[,1],main="eigenvalues",names.arg=1:nrow(res.pca$eig)) 
fviz_eig(res.pca, addlabels = TRUE)
#2 dimensions
```

#Individual result
```{r individual contribution and coordinates, results = "hide"}
res.pca$ind$contrib
res.pca$ind$coord
```

```{r individual plot}
fviz_pca_ind(res.pca, geom.ind="point",
             col.ind=inca_pca$bmi_bin,
             palette=c("blue","yellow"),
             legend.title="Nutritional Status",
             addEllipses = T,
             title="Individual Plot based on Nutritional Status")
```

#Variable result
```{r graph variable}
fviz_pca_var(res.pca, axes=c(1,2), col.var="contrib",
             gradient.cols=c("green","orange","pink"),
             repel=T,
             title="Correlation circle by distribution")
```


```{r variable contributions, results = "hide"}
res.pca$var
```

```{r supplementary only}
plot.PCA(res.pca, axes=c(1, 2), choix="ind", habillage = "none", invisible=c("ind"), new.plot=F, graph.type="classic")
```

##Clustering: HCA on PCA
#2 clusters
```{r perform HCA automatic}
res.pca.hcpc <- HCPC(res.pca , nb.clust=-1, consol=T, min=2, max=10, graph=TRUE)
```

```{r plot 2 cluster}
fviz_cluster(res.pca.hcpc)
```

#3 clusters
```{r perform HCA 3 clusters}
res.pca.hcpc3 <- HCPC(res.pca , nb.clust=3, consol=T, min=2, max=10, graph=TRUE)
```

```{r plot 3 cluster}
fviz_cluster(res.pca.hcpc3) #Choose 3 clusters
```

#Add cluster to dataset
```{r add cluster to dataset}
#3 clusters
inca.cluster.data <- cbind.data.frame(inca_clean, res.pca.hcpc3$data.clust$clust)
names(inca.cluster.data)[names(inca.cluster.data) == "res.pca.hcpc3$data.clust$clust"] <- "cluster3"
```

#describe cluster
```{r cluster 3 describe, results = "hide"}
res.pca.hcpc3$desc.var$quanti #clusters described by the quantitative active variables. 

res.pca.hcpc3$desc.var$category #clusters described by the illustrative qualitative variables.

res.pca.hcpc3$desc.var$test.chi2 #disease is significant, change outcome to disease
```


```{r cluster 3}
#check chi square between 
chisq.test(inca.cluster.data$bmi_bin, inca.cluster.data$cluster3) #p-value 4.45x10^-13
```

```{r descriptive graph}
inca_descriptive <- inca.cluster.data[,-c(4,8,9:50, 52)]
colnames(inca_descriptive) <-c('Diet','Chronic Disease','Season','Age category','Income Level','Physical Activity','Supplement Consumption','Household Category','Nutritional Status', 'Smoking','Education','Cluster')

inca_descriptive %>% tbl_summary(by = Cluster) %>% add_p() %>% modify_spanning_header(c("stat_1", "stat_2","stat_3") ~ "**Cluster**") %>% as_gt()
```

## Multivariate Regression

Run bivariate analysis between clusters, other confonding variables, against nutritional status.
```{r bivariate}
inca_regression <- inca_descriptive[,-c(1,3,7,8)]
inca_regression %>% tbl_summary(by = `Nutritional Status`) %>% add_p() %>% modify_spanning_header(c("stat_1", "stat_2") ~ "**Nutritional Status**") %>% as_gt()
```

```{r logistic regression}
inca_regression$`Chronic Disease` <- relevel(inca_regression$`Chronic Disease`, ref="No CD")
inca_model <- glm(data=inca_regression, `Nutritional Status`~`Cluster`+`Age category`+`Smoking`+`Chronic Disease`+`Education`+`Physical Activity`, family=binomial(link="logit"))
summary(inca_model)

tbl_regression(inca_model, exponentiate = T) %>% as_gt()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
