---
title: "Lab2_MCA"
format: html
editor: visual
---

# Load packages

```{r}
#| echo: true
#| eval: true
#| warning: false
#| message: false
#| 
library(FactoMineR)
library(factoextra)
library(foreign)

```

# Data importation

```{r}
#| echo: true
hemo <- read.table("hemo_quali.txt", header =T, sep = ",", dec = ".")

```

# Recoding variables as factors and adding labels to qualitatives variables

```{r}
#| echo: true
hemo$recodcir <- factor(hemo$recodcir, levels =c("1","2"), labels = c("cir_yes", "cir_no"))
hemo$alcoolisme <- factor(hemo$alcoolisme, levels =c("OH=1","OH=2"), labels = c("Alcohol_yes", "Alcohol_no"))
hemo$hepatomegalie <- factor(hemo$hepatomegalie)
hemo$insuffisance_hepato_cellulaire <- factor(hemo$insuffisance_hepato_cellulaire)

hemo$alt_en_classes <- factor(hemo$alt_en_classes, levels = c("<0.84",">=0.84"), labels = c("low alanine", "high alanine"))
hemo$ast_en_classes <- factor(hemo$ast_en_classes, levels = c("<0.67", ">=0.67"), labels = c("low aspart", "high aspart"))
hemo$ggt_en_classes <- factor(hemo$ggt_en_classes, levels = c("<0.98", ">=0.98"),labels = c("low ggt", "high ggt"))
hemo$bmi_en_classes <- factor(hemo$bmi_en_classes, levels = c("<23.84", ">=23.84"),labels = c("normal", "overweight"))
hemo$agediag_en_classes <- factor(hemo$agediag_en_classes, levels = c("<46", ">=46"),labels = c("young", "old"))
hemo$cs_en_classes <- factor(hemo$cs_en_classes, levels = c("<79", ">=79"),labels = c("Translow", "Transhigh"))
hemo$chfage_en_classes <- factor(hemo$chfage_en_classes, levels = c("<7.36", ">=7.36"),labels = c("hepaticlow", "hepatichigh"))
hemo$fer_en_classes <- factor(hemo$fer_en_classes, levels = c("<37", ">=37"),label = c("ironlow", "ironhigh"))
```

# Selection of active categorical variables and illustrative variables (recodcir)

Choose all variables, except "ident"

```{r}
hemo.MCA<-subset(hemo,select = -ident)

```

# Perform a MCA

Choose the diagnosis of cirrhosis as an illustrative variable. We can integrate all the other variables, including sex as active ones.

```{r}
res.mca<-MCA(hemo.MCA, ncp=3, quali.sup= 5,graph = F)

```

Observe the eigenvalues and the screeplot to choose the number of dimensions to interpret

```{r}
#| echo: true
#| #| warning: false
#| message: false
res.mca$eig

#the sum of eigen values must be = (K/J)-1
eigenvalue <-data.frame(round(res.mca$eig[,1],2)) 
sum.eig <- apply(eigenvalue, 2, sum)
sum.eig

fviz_screeplot(res.mca, addlabels = TRUE)
# select the number of component to be interpreted

```

# Display the graphic of individuals and variables

Biplot with factoextra = both individuals and categories of variables in the same graphic

```{r}
#| echo: true
#| #| warning: false
#| message: false
#| 

fviz_mca_biplot(res.mca)
# use repel= TRUE option to avoid text overlapping (slow if many point)

```

Graphic only with categories of variables

```{r}
#| echo: true
#| #| warning: false
#| message: false

#graphic with categories of variables only colored following their coordinates to dim 1

fviz_mca_var(res.mca, col.var = "coord",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE, # Avoid text overlapping
             ggtheme = theme_minimal())

#graphic with categories of variables only colored following their contribution to dim 1

fviz_mca_var(res.mca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE, # Avoid text overlapping
             ggtheme = theme_minimal())
```

Graphic only with individuals

```{r}

#graphic with individuals only colored following the illustrative variable "recodcir"
#| echo: true
#| #| warning: false
#| message: false
#| 
fviz_mca_ind(res.mca,
             geom.ind = "point", #point not number of individuals 
             col.ind = hemo$recodcir, # color according to diagnosis of cirrhosis
             palette = c("#00AFBB", "#E7B800"),
             legend.title = "Cirrhosis",
             addEllipses = T, # Concentration ellipses
             title="Graphic of individuals following a diagnosis of cirrhosis"
)
```

# Numerical outputs of individuals and variables

In MCA, beyond contributions and coordinates, another indicator is available : the v-test parameter discriminates the categories of variables that are significantly contributing to the inertia of the dimension. its critical value is 2.0 (negative or positive). you should discard the contributions of the original variables that do not reach this value

```{r}
#the results of the individuals are available using the following commands
res.mca$ind #present all indicators
res.mca$ind$contrib #present only contributions
res.mca$ind$coord #present only coordinates

#the results of the categories of variables are available using the following commands
res.mca$var  #present all indicators 
res.mca$var$v.test #present only v-test
res.mca$var$contrib #present only contributions
res.mca$var$coord #present only coordinates

```

# Use explor to show the results of the MCA through an interactive platform

```{r}

library(explor)
explor(res.mca)

```
